---
title: "Creating Data Packages in R"
author: "Kleanthis Koupidis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This tutorial will show you how to install the R library for working with Data Packages and Table Schema, load a CSV file, infer its schema, and write a Tabular Data Package.

#Setup

For this tutorial, we will need the Data Package R library ([datapackage.r](https://github.com/frictionlessdata/datapackage-r)).

[devtools library](https://cran.r-project.org/package=devtools) is required to install the datapackage.r library from github.

```{r, eval=FALSE, include=TRUE}
# Install devtools package if not already
install.packages("devtools")
```

And then install the development version of [datapackage.r](https://github.com/frictionlessdata/datapackage-r) from github.

```{r, eval=FALSE, include=TRUE}
devtools::install_github("frictionlessdata/datapackage.r")
```

#Load

You can start using the library by loading `datapackage.r`.

```{r, eval=TRUE, include=TRUE}
library(datapackage.r)
```

You can add useful metadata by adding keys to metadata dict attribute. Below, we are adding the required `name` key as well as a human-readable `title` key. For the keys supported, please consult the full [Data Package spec](https://frictionlessdata.io/specs/data-package/#metadata). Note, we will be creating the required `resources` key further down below.

```{r, eval=TRUE, include=TRUE}
dataPackage = Package.load()
dataPackage$descriptor['name'] = 'period-table'
dataPackage$descriptor['title'] = 'Periodic Table'
dataPackage$commit()
```

#Infer a CSV Schema

We will use periodic-table data from remote path: <https://github.com/frictionlessdata/example-data-packages/blob/master/periodic-table/data.csv>

We can guess at our CSV's [schema](https://frictionlessdata.io/guides/table-schema/) by using `infer` from the Table Schema library. We pass directly the remote link to the infer function, the result of which is an inferred schema. For example, if the processor detects only integers in a given column, it will assign `integer` as a column type.

```{r, eval=TRUE, include=TRUE}
filepath = 'https://raw.githubusercontent.com/frictionlessdata/example-data-packages/master/periodic-table/data.csv'

schema = tableschema.r::infer(filepath)
```

Once we have a schema, we are now ready to add a `resource` key to the Data Package which points to the resource path and its newly created schema. Below we define resources with two ways, using json text format and list objects.

```{r, eval=TRUE, include=TRUE}
# define resources using json text 
resources = helpers.from.json.to.list(
  '[{
	"name": "data",
	"path": "filepath",
	"schema": "schema"
  }]'
)
resources[[1]]$schema = schema
resources[[1]]$path = filepath

# define resources using list object
resources = list(
  list( name = "data",
  path = filepath,
  schema = schema
  ))

```

And now, add resources to the Data Package:

```{r, eval=TRUE, include=TRUE}
dataPackage$descriptor['resources'] = resources
```

Now we are ready to write our `datapackage.json` file to the current working directory.

```{r, eval=FALSE, include=TRUE}
dataPackage$save(getwd())
```

The `datapackage.json` ([download](https://github.com/frictionlessdata/example-data-packages/blob/master/periodic-table/datapackage.json)) is inlined below. Note that atomic number has been correctly inferred as an `integer` and atomic mass as a `number` (float) while every other column is a `string`.

```{r, eval=TRUE, include=TRUE}
jsonlite::toJSON(dataPackage$descriptor, pretty = TRUE)
```

#Publishing

Now that you have created your Data Package, you might want to [publish your data online](https://frictionlessdata.io/guides/publish-online/) so that you can share it with others.