---
title: "Using Data Packages in R"
author: "Kleanthis Koupidis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Data Packages in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This tutorial will show you how to install the R libraries for working with Tabular Data Packages and demonstrate a very simple example of loading a Tabular Data Package from the web and pushing it directly into a local SQL database. Short examples of pushing your dataset to Google’s BigQuery and Amazon’s RedShift follow.

#Setup

For this tutorial, we will need the Data Package R library ([datapackage.r](https://github.com/frictionlessdata/datapackage-r)).

[devtools library](https://cran.r-project.org/package=devtools) is required to install the datapackage.r library from github.

```{r, eval=FALSE, include=TRUE}
# Install devtools package if not already
install.packages("devtools")
```

And then install the development version of [datapackage.r](https://github.com/frictionlessdata/datapackage-r) from github.

```{r, eval=FALSE, include=TRUE}
devtools::install_github("frictionlessdata/datapackage.r")
```

#Load

You can start using the library by loading `datapackage.r`.

```{r, eval=TRUE, include=TRUE}
library(datapackage.r)
```

#Reading Basic Metadata

In this case, we are using an example Tabular Data Package containing the periodic table stored on [GitHub](https://github.com/frictionlessdata/example-data-packages/tree/master/periodic-table) ([datapackage.json](https://raw.githubusercontent.com/frictionlessdata/example-data-packages/master/periodic-table/datapackage.json), [data.csv](https://raw.githubusercontent.com/frictionlessdata/example-data-packages/master/periodic-table/data.csv)). This dataset includes the atomic number, symbol, element name, atomic mass, and the metallicity of the element. Here are the first five rows:


```{r, echo=FALSE, results='asis'}
url = 'https://raw.githubusercontent.com/okgreece/datapackage-r/master/vignettes/example_data/data.csv'
pt_data = read.csv2(url, sep = ',')
knitr::kable(head(pt_data, 5), align = 'c')
```

Data Packages can be loaded either from a local path or directly from the web.

```{r}
url = helpers.from.json.to.list('https://raw.githubusercontent.com/okgreece/datapackage-r/master/vignettes/example_data/package.json')
datapackage = Package.load(url)
# resource = Resource.load(url,basePath = 'https://raw.githubusercontent.com/frictionlessdata/example-data-packages/master/periodic-table' )
# resource$read()
```

At the most basic level, Data Packages provide a standardized format for general metadata (for example, the dataset title, source, author, and/or description) about your dataset. Now that you have loaded this Data Package, you have access to this `metadata` using the metadata dict attribute. Note that these fields are optional and may not be specified for all Data Packages. For more information on which fields are supported, see [the full Data Package standard](https://frictionlessdata.io/specs/data-package/).

```{r}
# datapackage$resources[[1]]$read(keyed= TRUE)
```

#Reading Data

Now that you have loaded your Data Package, you can read its data. A Data Package can contain multiple files which are accessible via the `resources` attribute. The `resources` attribute is an array of objects containing information (e.g. path, schema, description) about each file in the package.

You can access the data in a given resource in the `resources` array by reading the `data` attribute. For example, using our our Periodic Table Data Package, we can return all elements with an atomic number of less than 10 by doing the following:

print( 
[
e['name'] for e in dp.resources[0].data if int(e['atomic number']) < 10
]
)

```{r}
datapackage$descriptor$resources[[1]]$path
def = tableschema.r::Table.load('https://raw.githubusercontent.com/okgreece/datapackage-r/master/vignettes/example_data/data.csv')
table=def$value()
periodic_table_data = table$read(keyed = TRUE)
str(periodic_table_data)
```


