---
title: "Using Data Packages in R"
author: "Kleanthis Koupidis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Data Packages in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This tutorial will show you how to install the R packages for working with Tabular Data Packages and demonstrate a very simple example of loading a Tabular Data Package from the web and pushing it directly into a local SQL database and send query to retrieve results.

# Setup

For this tutorial, we will need the Data Package R package ([datapackage.r](https://github.com/frictionlessdata/datapackage-r)).

[devtools package](https://cran.r-project.org/package=devtools) is required to install the datapackage.r package from github.

```{r, eval=FALSE, include=TRUE}
# Install devtools package if not already
install.packages("devtools")
```

And then install the development version of [datapackage.r](https://github.com/frictionlessdata/datapackage-r) from github.

```{r, eval=FALSE, include=TRUE}
install.packages("datapackage.r")
# or install the development package
devtools::install_github("frictionlessdata/datapackage.r")
```

# Load

You can start using the package by loading `datapackage.r`.

```{r, eval=TRUE, include=TRUE}
library(datapackage.r)
```

# Reading Basic Metadata

In this case, we are using an example Tabular Data Package containing the periodic table stored on [GitHub](https://github.com/frictionlessdata/examples/tree/master/periodic-table) ([datapackage.json](https://raw.githubusercontent.com/frictionlessdata/example-data-packages/master/periodic-table/datapackage.json), [data.csv](https://raw.githubusercontent.com/frictionlessdata/example-data-packages/master/periodic-table/data.csv)). This dataset includes the atomic number, symbol, element name, atomic mass, and the metallicity of the element. Here are the first five rows:


```{r, echo=TRUE, results='asis'}
path <- 'exampledata/data.csv' # or use url <- 'https://raw.githubusercontent.com/frictionlessdata/datapackage-r/master/vignettes/exampledata/data.csv'
pt_data <- read.csv2(path, sep = ',')
knitr::kable(head(pt_data, 5), align = 'c')
```

Data Packages can be loaded either from a local path or directly from the web.

```{r, eval=TRUE, include=TRUE}
path <- 'exampledata/package.json' # or use url <- 'https://raw.githubusercontent.com/frictionlessdata/datapackage-r/master/vignettes/exampledata/package.json'
datapackage <- Package.load(path)
datapackage$resources[[1]]$descriptor$profile <- 'tabular-data-resource' # tabular resource descriptor profile 
datapackage$resources[[1]]$commit() # commit changes
```

At the most basic level, Data Packages provide a standardized format for general metadata (for example, the dataset title, source, author, and/or description) about your dataset. Now that you have loaded this Data Package, you have access to this `metadata` using the metadata dict attribute. Note that these fields are optional and may not be specified for all Data Packages. For more information on which fields are supported, see [the full Data Package standard](https://specs.frictionlessdata.io//data-package/).

```{r, eval=TRUE, include=TRUE}
datapackage$descriptor$title
```

# Reading Data

Now that you have loaded your Data Package, you can read its data. A Data Package can contain multiple files which are accessible via the `resources` attribute. The `resources` attribute is an array of objects containing information (e.g. path, schema, description) about each file in the package.

You can access the data in a given resource in the `resources` array by reading the `data` attribute. 

```{r, eval=TRUE, include=TRUE}
table <- datapackage$resources[[1]]$table
periodic_table_data <- table$read()
```

You can further manipulate list objects in R by using [purrr](https://cran.r-project.org/package=purrr), [rlist](https://cran.r-project.org/package=rlist) packages.

# Loading into an SQL database

[Tabular Data Packages](https://specs.frictionlessdata.io//tabular-data-package/) contains schema information about its data using [Table Schema](https://specs.frictionlessdata.io//table-schema/). This means you can easily import your Data Package into the SQL backend of your choice. In this case, we are creating an [SQLite](http://sqlite.org/) database.

To create a new SQLite database and load the data into SQL we will need [DBI](https://cran.r-project.org/package=DBI) package and [RSQLite](https://cran.r-project.org/package=RSQLite) package, which contains [SQLite](https://www.sqlite.org/index.html) (no external software is needed).

You can install and load them by using:

```{r, eval=FALSE, include=TRUE}
install.packages(c("DBI","RSQLite"))
```

```{r, eval=TRUE, include=TRUE}
library(DBI)
library(RSQLite)
```

To create a new SQLite database, you simply supply the filename to `dbConnect()`:

```{r, eval=TRUE, include=TRUE}
dp.database <- dbConnect(RSQLite::SQLite(), "") # temporary database
```

We will use data.table package to convert the list object with the data to a data frame in order to copy them to database table.

```{r, eval=TRUE, include=TRUE}
# install data.table package if not already
# install.packages("data.table")

periodic_table_sql <- data.table::rbindlist(periodic_table_data)
periodic_table_sql <- setNames(periodic_table_sql,unlist(datapackage$resources[[1]]$headers))
```

You can easily copy an R data frame into a SQLite database with dbWriteTable():

```{r, eval=TRUE, include=TRUE}
dbWriteTable(dp.database, "periodic_table_sql", periodic_table_sql)
# show remote tables accessible through this connection
dbListTables(dp.database)
```
The data are already to the database.

We can further issue queries to hte database:

Return first 5 elements:
```{r, eval=TRUE, include=TRUE}
dbGetQuery(dp.database, 'SELECT * FROM periodic_table_sql LIMIT 5')
```

Return all elements with an atomic number of less than 10:

```{r, eval=TRUE, include=TRUE}
dbGetQuery(dp.database, 'SELECT * FROM periodic_table_sql WHERE "atomic number" < 10')
```

More about using databases, SQLite in R you can find in vignettes of [DBI](https://cran.r-project.org/package=DBI) and [SQLite](https://www.sqlite.org/index.html) packages.
